Testing Efficacy; Insights from a Simple Epidemic Model

### 2020-11-15 SIR_manuscript.tex
Add files via upload

@@

@@

- "delay negative results" result

>>>>

%%Packages and commands added by Fady
\usepackage{xcolor}
\newcommand{\fady}[1]{\textcolor{cyan}{$\langle${\slshape{\bfseries Fady:} #1 }$\rangle$}}
\newcommand{\Rnum}{\mathcal{R}_0}
\usepackage{xspace}
\usepackage[nameinlink,capitalize]{cleveref}

@@

\subsection{rate of returning tests}
The linearization of $\Rnum$ around $\rho=0$ is
\begin{equation}\label{linearization}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 W_s} \Big(\gamma(\eta_w-1)(\gamma W_s+\omega W_i) + (\eta_t -1)P_iW_i \omega^2 \Big). 
\end{equation}

So when $\rho \approx 0$ we have $$\partial{\Rnum}/\partial{\omega} \approx  \frac{-\beta \rho}{\gamma W_s\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c).$$

where $a=(\eta_w-1)W_i-(\eta_t-1)P_iW_i$, $b=2(\eta_w-1)\gamma W_s$ and $c=(\eta_w-1)\gamma^2 W_s$.

Perhaps counter-intuitively, the equation above does not predict that $\Rnum$ is monotone decreasing with respect to $\omega$. In other words; our model does not predict that returning test results more rapidly \textit{always} lower $\Rnum$. In order to gain insight into this intriguing behaviour, we examine the zeroes of $\frac{\partial{\Rnum}}{\partial{\omega}}(\omega)$.

Defining the following quantity, $Q$, will help us write the roots of $\partial{\Rnum}/\partial{\omega}$ neatly. 
\begin{align}\label{eq:defQ}
    Q =& \frac{W_i}{W_s}\left(1-\frac{n_{t}-1}{n_{w}-1}P_{i}\right) \\
\end{align}

With that in mind, we can write the roots of $\partial{\Rnum}/\partial{\omega}$ as

\begin{align}
    \omega_1 =& \frac{\gamma}{-\sqrt{1-Q}-1} \\
    \omega_2 =& \frac{\gamma}{\sqrt{1-Q}-1}
\end{align}

Note that the zeroes are real if and only if $Q < 1$. Note that have $\eta_t < \eta_w$, so if $P_i \approx 1$, we will have $Q < 0 < 1$. Thus, if we assume near-perfect test sensitivity, $\omega_1$ and $\omega_2$ will be real.

Assuming $\omega_1, \omega_2$ are real, it is easy to confirm that $\omega_1 < 0$ by looking at the denominator. To see that $\omega_2 > 0$, recall that $Q < 0$, so $\sqrt{1-Q} > 1$ and so $\sqrt{1-Q} -1 > 0$. Knowing that $\omega_1 < 0$, the only root of interest (i.e., biologically relevant quantity) is $\omega_2$.

We can prove that $\partial{\Rnum}/\partial{\omega} > 0$ when $\omega \in (0,\omega_2)$ and $\partial{\Rnum}/\partial{\omega} < 0$ when $\omega \in (\omega_2,\infty)$ by computing the limits of $\partial{\Rnum}/\partial{\omega}$ at $0$ and $\infty$ respectively. So it follows that $\Rnum$ has a global maximum with respect to $\omega$ at $\omega = \omega_2$.

Now we want to characterize the parameter regions on which $\partial{\Rnum}/\partial{\omega} < 0$ (i.e., the conditions under which returning test results more rapidly is favourable). By the previous analysis, this is equivalent to solving for $\omega > \omega_2$. So

\begin{align}\label{eq:necsuf}
    &\omega > \omega_2 \nonumber \\
    &\omega > \frac{\gamma}{\sqrt{1-Q}-1} \nonumber \\
    &\vdots \nonumber \\
    &\frac{1-n_{t}}{1-n_{w}}P_{i}>\frac{W_{s}}{W_{i}}\left(\frac{\gamma}{\omega}+1\right)^{2}
\end{align}

This inequality quantifies the exact relationships between model parameters that result in returning tests more rapidly being favourable. To give a biological interpretation, we describe the qualitative \textit{trends} predicted by the inequality. Returning test results more rapidly \textit{tends} to be more favourable when

\begin{itemize}
    \item individuals who have tested positive lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_t \ll \eta_w$).
    \item the test is sensitive.
    \item testing of infected individuals is more intense than testing of susceptible individuals.
\end{itemize}
An in-depth discussion of these results is presented in the discussion section.

\subsection{Expensive vs. cheap tests}

The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. In the analysis below, we compare the $\Rnum$ predicted by our model depending on the testing strategy.

Consider a test that allows us to test at rate $\rho_1$ and has sensitivity $P_{i,1}$, and another test that allows us to test at $\rho_2$ and has sensitivity $P_{i,2}$. Suppose that $\rho_1 > \rho_2$. Recall that the linearization of $\Rnum$ around $\rho \approx 0$ is given by $$\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 W_s} \Big(\gamma(\eta_w-1)(\gamma W_s+\omega W_i) + (\eta_t -1)P_iW_i \omega^2 \Big).$$

Treating $\Rnum$ as a function of $\rho$ and $P_i$,we can reduce the inequality $$\Rnum(\rho_2, P_{i,2}) < \Rnum(\rho_1, P_{i,1})$$ into

\begin{align}\label{eq:rho1vsrho2}
    &\rho_1\left(\gamma(\eta_w-1)(\gamma W_s + \omega W_i) + (\eta_t-1)P_{i, 1}W_i\omega^2\right) - \rho_2\left(\gamma(\eta_w-1)(\gamma W_s + \omega W_i) + (\eta_t-1)P_{i, 2}W_i\omega^2\right) > 0 \nonumber \\
    &\vdots \nonumber \\
    &\frac{\rho_2P_{i, 2}-\rho_1P_{i, 1} }{\rho_1-\rho_2} > \frac{1-\eta_w}{1-\eta_t}\cdot \frac{\gamma(\gamma W_s + \omega W_i)}{\omega^2 W_i}
\end{align}

Note that the RHS is positive, thus a necessary condition for the inequality above to hold is that $\rho_2P_{i,2} > \rho_1P_{i,1}$, equivalently

\begin{equation}
\frac{P_{i,2}}{P_{i,1}} > \frac{\rho_1}{\rho_2}.
\end{equation}

To state an example of this, if test $A$ is three times as expensive as test $B$ (and hence one can test three times as many people with test $B$), using test $A$ rather than $B$ will be favourable only if test $A$ is at least 3 times more sensitive than test $B$. Note that this is a necessary but not sufficient condition, so even if test $A$ is three times more sensitive, it is still possible for test $B$ to be more effective.

\cref{eq:rho1vsrho2} tells us precisely when a test corresponding to $\rho_2, P_{i,2}$ will yield a lower $\Rnum$ than a test corresponding to $\rho_1, P_{i,1}$, where $\rho_1 > \rho_2$. Some of the qualitative \textit{trends} that favour test 2 (the higher-sensitivity test) include

\begin{itemize}
    \item individuals who test positive self-isolate much more than individuals who are waiting for their test result.
    \item the time it takes to return tests is much shorter than the mean infectious period.
    \item the testing intensity is much greater for infected individuals than susceptible individuals.
\end{itemize}

@@

\subsection{Rate of returning tests}
\fady{needs to be expanded}
When the difference in self-isolation between individuals waiting for their tests and individuals who are confirmed positive is small, returning tests rapidly is not beneficial because the benefit in doing so (i.e., the extra self-isolating when an individual tests positive) is minimal and outweighed by the fact that those who test negative may increase their contacts. Similarly, if the test is not sensitive, many infected individuals will test negative, and thus may increase their contacts under the false belief that they are not infectious, hence returning test results more quickly in this case may not be beneficial.

When testing of infected individuals is more intense than susceptible individuals, returning test results rapidly is beneficial because those infected individuals will be informed and self-isolate. Conversely, if testing of susceptible individuals is much greater, very few tests be positive, and the benefit of informing those few positive cases will be outweighed by the potential increase in contacts of the individuals testing negative.

\subsection{Expensive vs. cheap tests}
\fady{needs to be written \dots}
